---
title: "figures?"
output: html_document
date: "2023-11-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(arrow)
library(lubridate)
library(tidytext)
```

## Read in the data and model metadata

```{r read-in-data}

lake_sites <- c('BARC', 'CRAM', 'LIRO', 'PRLA', 'PRPO', 'SUGG', 'TOOK')
cols_modeltype <- c('empirical' = "#482576FF",
          'multi-model ensemble' =  "#35608DFF",
          'process' = "#21908CFF",
          'ML' = '#BBDF27FF',
          'null' = '#43BF71FF')
# read in the metadata from the google sheet
googlesheets4::gs4_deauth()
model_meta <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1oC7_w63wSCXNiHs1IK8AFGr0MG-NdjDAjwkfjvRZW-I/edit?usp=sharing")

# Open dataset-------------
# start with lake temperatures
temperature_scores <- arrow::open_dataset("../scores_rescored") |>
  filter(site_id %in% lake_sites,
         # model_id != 'climatology',
         model_id %in% model_meta$model_id,
         variable == 'temperature') |>
  collect()


# Calculate skill scores ------------
temperature_climatology <- arrow::open_dataset("../scores_rescored") |>
  filter(model_id == 'climatology', 
         site_id %in% lake_sites,
         variable == 'temperature') |>
  collect() |> 
  select(reference_datetime, datetime, crps, logs, site_id, model_id) |> 
  arrange(reference_datetime, datetime, site_id) |> 
  pivot_wider(values_from = c(crps, logs),
              names_from = model_id)

temperature_skill <- temperature_scores |>
  filter(model_id != 'climatology') |> 
  select(reference_datetime, datetime, crps, logs, site_id, model_id) |> 
  full_join(temperature_climatology) |> 
  
  # calcaulte the skill relative to climatology (- is bad, + is good)
  mutate(skill_crps = crps_climatology - crps,
         skill_logs = logs_climatology - logs,
         horizon = as.numeric(as_date(datetime) - as_date(reference_datetime))) |> 
  # consistent forecast period
  filter(horizon <= 30,
         horizon >= 1,
         model_id != 'fARIMA') |> 
  arrange(reference_datetime, site_id, datetime, model_id) 
```

## Figure 1

Model type impacts of water temperature forecast skill (relative to climatology). Individual model mean CRPS skill, coloured by model_type (as defined by submitters). 
Overall and then by site. 
```{r model-type, fig.width=10}
top5_temp_site <- temperature_skill |> 
  group_by(model_id, site_id) |> 
  summarise(mean_crps = mean(skill_crps, na.rm = T)) |> 
  group_by(site_id) |> 
  slice_max(mean_crps, n= 5, na_rm = T) |> 
  mutate(model_id = fct_reorder(model_id, desc(mean_crps))) |>
  left_join(model_meta) |> 
  ggplot(aes(x=mean_crps, 
             y = reorder_within(x = model_id, within = site_id, by = mean_crps),
             fill = model_type)) +
  geom_bar(stat = 'identity') +
  scale_y_reordered() +
  scale_fill_manual(values = cols_modeltype) +
  facet_wrap(~site_id, scales = 'free_y',nrow = 4) +
  labs(y='model_id', x = 'mean CRPS skill (vs climatology)') +
  geom_vline(xintercept = 0) +
  theme_bw()


top10_temp <- temperature_skill |> 
  group_by(model_id) |> 
  summarise(mean_crps = mean(skill_crps, na.rm = T)) |> 
  ungroup() |> 
  slice_max(mean_crps, n= 10, na_rm = T) |> 
  left_join(model_meta) |> 
  mutate(model_id = fct_reorder(model_id, mean_crps)) |>
  ggplot(aes(x=mean_crps, 
             y = model_id,
             fill = model_type)) +
  geom_bar(stat = 'identity') +
  scale_y_reordered() +
  labs(y='Model ID', x = 'mean CRPS skill (vs climatology)') +
  geom_vline(xintercept = 0) +
  scale_fill_manual(values = cols_modeltype) +
  theme_bw()

ggpubr::ggarrange(top10_temp, top5_temp_site, nrow = 1, 
                  # widths = c(1,0.5), 
                  common.legend = T,
                  legend = 'top')

```
 
## Figure 2
How does model type performance, on avergae, differ by site. Either comparing sites by model type or vise versa. 
This uses all model_ids (maybe get rid of the 'worst') performers. 

```{r model_type-site}
ggpubr::ggarrange(temperature_scores |> 
  left_join(model_meta, by = 'model_id') |> 
  group_by(model_type, model_id, site_id) |> 
  summarise(mean_crps = mean(crps, na.rm = T)) |>  
  ggplot(aes(x = mean_crps, 
             y = site_id)) +
  geom_boxplot() +
  facet_wrap(~model_type, scales = 'free_y',nrow = 4) +
  labs(y='model_id', x = 'mean CRPS score') +
  theme_bw(),
  
  temperature_scores |> 
  left_join(model_meta, by = 'model_id') |> 
  group_by(model_type, model_id, site_id) |> 
  summarise(mean_crps = mean(crps, na.rm = T)) |>  
  ggplot(aes(x = mean_crps, 
             y = model_type)) +
  geom_boxplot() +
  facet_wrap(~site_id, scales = 'free_y',nrow = 4) +
  labs(y='model_id', x = 'mean CRPS score') +
  theme_bw(),
  nrow = 1, widths = c(1, 1.5))

```

## Figure 3
How does the performance vary across horizon? This is the top 10 models at each site (so this doesn't really look at model type...)

```{r horizon-score}
# top 10 average models
top_mods <- temperature_skill |> 
  group_by(model_id) |> 
  summarise(mean_crps = mean(skill_crps, na.rm = T)) |> 
  ungroup() |> 
  slice_max(mean_crps, n = 10, na_rm = T) |> 
  distinct(model_id)

temperature_skill |> 
  filter(model_id %in% top_mods$model_id) |> 
  group_by(site_id, horizon) |> 
  summarise(mean_crps = mean(skill_crps, na.rm = T)) |> 
  ggplot(aes(x=horizon, 
             y = mean_crps, colour = site_id)) +
  geom_hline(yintercept = 0) + 
  geom_line()  +  
  labs(y='mean CRPS skill (vs climatology)', x = 'horizon (days)') +
  theme_bw()

```

So maybe something more like this...
This is all sites together, so more like figure 1.

```{r NOAA-scores}
temperature_skill |> 
  filter(model_id %in% top_mods$model_id) |> 
  group_by(model_id, horizon) |> 
  summarise(mean_crps = mean(skill_crps, na.rm = T)) |> 
  ggplot(aes(x=horizon, 
             y = mean_crps, colour = model_id)) +
  geom_hline(yintercept = 0) + 
  geom_line()  +  
  labs(y='mean CRPS skill (vs climatology)', x = 'horizon (days)') +
  theme_bw()
```

## Figure 4 - inclusion of covariates

```{r}

temperature_skill |> 
  group_by(model_id) |> 
  summarise(mean_crps = mean(skill_crps, na.rm = T)) |> 
  ungroup() |> 
  slice_max(mean_crps, n= 10, na_rm = T) |> 
  left_join(model_meta) |> 
  mutate(model_id = fct_reorder(model_id, mean_crps)) |>
  ggplot(aes(x=mean_crps, 
             y = model_id,
             fill = as.factor(uses_NOAA))) +
  geom_bar(stat = 'identity') +
  scale_y_reordered() +
  labs(y='Model ID', x = 'mean CRPS skill (vs climatology)') +
  geom_vline(xintercept = 0) +
  scale_fill_viridis_d(option = 'inferno') +
  theme_bw()

temperature_skill |> 
  group_by(model_id, site_id) |> 
  summarise(mean_crps = mean(skill_crps, na.rm = T)) |> 
  group_by(site_id) |> 
  slice_max(mean_crps, n= 5, na_rm = T) |> 
  mutate(model_id = fct_reorder(model_id, desc(mean_crps))) |>
  left_join(model_meta) |> 
  ggplot(aes(x=mean_crps, 
             y = reorder_within(x = model_id, within = site_id, by = mean_crps),
             fill = as.factor(uses_NOAA))) +
  geom_bar(stat = 'identity') +
  scale_y_reordered() +
  scale_fill_viridis_d(option = 'inferno') +
  facet_wrap(~site_id, scales = 'free_y',nrow = 4) +
  labs(y='model_id', x = 'mean CRPS skill (vs climatology)') +
  geom_vline(xintercept = 0) +
  theme_bw()
```


## Figure 5 - NOAA scores

can this be related to the weather forecasts?

```{r}
NOAA_scores <- arrow::open_dataset("../scores") |>
  filter(site_id %in% lake_sites,
         model_id %in% c('NOAA_daily', 'NOAA_daily_obs')) |>
  collect()

NOAA_scores |> 
  mutate(horizon = as_date(datetime) - as_date(reference_datetime)) |> 
  group_by(model_id, site_id, horizon) |> 
  summarise(mean_crps = mean(crps, na.rm = T)) |> 
  filter(horizon <= 30) |> 
  ggplot(aes(x=horizon, y=mean_crps, colour = model_id)) +
  geom_line() +
  facet_wrap(~site_id) +
  theme_bw() +
  labs(y = 'CRPS score (oC)', title = 'NOAA air temperature forecast')


```

## site differences in observations?

```{r obs-cv}
temperature_scores |> 
  mutate(horizon = as.numeric(as_date(datetime) - as_date(reference_datetime))) |> 
  filter(between(horizon, 0,30)) |> 
  group_by(site_id, variable, horizon) |> 
  summarise(cv = sd(observation, na.rm = T)/ mean(observation, na.rm = T)) |> 
  ggplot(aes(x=horizon, y= cv, colour = site_id)) +
  geom_point() +
  scale_colour_viridis_d(option = 'turbo', begin = 0.1, end = 0.9) +
  theme_bw()
```


Model rankings
```{r eval = F}
temperature_scores |> 
  mutate(horizon = as_date(datetime) - as_date(reference_datetime)) |> 
  filter(horizon < 30, horizon > 0) |> 
  select(reference_datetime, horizon, model_id, site_id, crps) |>
  arrange(reference_datetime, horizon, site_id, crps) |>
  group_by(reference_datetime, horizon, site_id) |>
  mutate(rank = row_number()) |>
  group_by(horizon, site_id, rank, model_id) |>
  summarise(proportion = (n()/length(unique(temperature_scores$reference_datetime)))) 

```

